---
title: Analysis of ddPCR data to confirm allele specific binding
author: Peter Humburg
date: `r format(Sys.time(), "%a %d %b %Y")`
---

```{r setup, include=FALSE}
library(knitr)
library(pander)
library(ggplot2)

opts_knit$set(root.dir = "/well/jknight/vivek")
opts_chunk$set(tidy=TRUE)
opts_chunk$set(echo=FALSE)
opts_chunk$set(autodep=TRUE)
options(figcap.prefix = "Figure", figcap.sep = ":", figcap.prefix.highlight = "**")
options(tabcap.prefix = "Table", tabcap.sep = ":", tabcap.prefix.highlight = "**")
```

# Introduction

```{r loadData}
data <- read.table("vivek_no_more_rain.txt", head=TRUE, sep="\t", check.names=FALSE)
data$Type <- factor(sapply(strsplit(as.character(data$Sample), " "), "[[", 1))
data$Sample <- factor(sapply(strsplit(as.character(data$Sample), " "), "[[", 2))
data$Allele <- factor(sub("NOD2", "", as.character(data$Assay)))
```
This is an analysis of ChIP ddPCR data to test whether there is allele specific binding 
to a region of the NOD2 gene. Two technical replicates of ChIP and input were generated
for `r length(levels(data$Sample))` samples. Measurements for the two 
alleles (`r pander(levels(data$Type))`) were taken from the same well, i.e. for each
replicate the two alleles are paired.

# Testing for allele specific binding
The counts obtained from the ddPCR measurements are assumed to come from a Poisson
distribution and concentration estimates are based on this. The concentration
estimates essentially are estimates of Poisson means that have been scaled to correspond
to the same unit volume ($1\muL$). Here we treat the concentration estimates as
Poisson means and test for a difference in the ratio between means for the two 
alleles^[Note that the estimates based on the raw data differ from the concentration 
estimates by approx. a factor of 1000].






# Apendix {-}
## Options
```{r setup, eval=FALSE}
```
## Session info
```{r sessionInfo}
sessionInfo()
```